---
title: "abruptShift_tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{abruptShift_tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(actR)
```


This notebook implements a test of the `actR::detectShiftCore()` function on synthetic data (see data_raw/makeSyntheticData.R for their construction). 

The idea is to use the simplest model of a shift (linear ramp emulating either a step or gradual change) to evaluate the proportion of false negatives and false positives: how often is a gradual shift mistaken for an abrupt one, or missed altogether? How often is a shift identified when none exists? 
From these rates we construct receiver operating characteristic curves and their sensitivity to:
- abruptness (the time over which the shift occurs, from 0 to T/2, where T is the length of the series)
- noise levels. 
- age uncertainties.

```{r Load test data }
lngth <- 400
time <- 1.0*seq(lngth)
pureshift <- linearRamp(lngth,width = 0)
ShiftHappens <- detectShift(time = time,
                            vals = pureshift,
                            vals.variable.name = 'Step Change',
                            time.units = 'years',
                            vals.units = 'unitless',
                            summary.bin.step = 1,
                            simulate.time.uncertainty = FALSE,
                            simulate.paleo.uncertainty = FALSE,
                            null.hypothesis.n = 50)

plot(ShiftHappens)
```


Now we'll do the same, by looping over various values of the shift


```{r gradual vs abrupt}
widths <- c(0, lngth/40, lngth/20, lngth/10, lngth/5, lngth/2)
rampRes = vector(mode = "list",length = length(widths))
for(w in 1:length(widths)) {
  rampRes[[w]] <- detectShift(time = time,
                            vals = linearRamp(lngth,width = widths[w]),
                            vals.variable.name = 'Ramp',
                            time.units = 'years',
                            vals.units = 'unitless',
                            summary.bin.step = 1,
                            simulate.time.uncertainty = FALSE,
                            simulate.paleo.uncertainty = FALSE,
                            null.hypothesis.n = 50)
  plot(rampRes[[w]])
}
```


