---
title: "Synthetic Data in  detectExcursion"
author: "Dave Edge and Nick McKay"
date: "2022-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```


# Synthetic excursions

## Build a synthetic time series with an excursion

Let's create a synthetic excursion centered on the 4.2 event
```{r}
syntheticExcursion <- makeExcursion(time = seq(3000,5000,by = 20), #specify the time
                                     amplitude = 1, #the amplitude of the excursion
                                     delta.amplitude = 0, #the change in amplitude, for asymmetrical excursions
                                     start.time = 4100, #when does the excursion start
                                     duration=200, #how long is the excusion
                                     snr = 1, #the signal to noise ratio
                                    time.variable.name = "Age",
                                    time.units = "yr BP") #signal to noise ratio.
                                    
```

Now, we can use `detectExcursion()` to test it. We'll plot the results after that.

## Detect Excursion

In `detectExcursion()` we need to specify a number of parameters. Check out the detectExcursion vignette for more details on how this works.

First, let's just test the excursion with now uncertainty or null hypothesis testing.

```{r}
detectedExcusion <- detectExcursion(syntheticExcursion,
                          event.yr = 4200,
                          event.window = 400,
                          ref.window = 400,
                          sig.num = 2,
                          exc.type = "positive",
                          n.ens = 1,
                          null.hypothesis.n = 1,
                          simulate.time.uncertainty = FALSE,
                          simulate.paleo.uncertainty = FALSE)
```

Now we can use `print()` and plot to check out the results

```{r}
print(detectedExcusion)
```

Let's plot the results

```{r}
plot(detectedExcusion)
```


## Excursion testing with uncertainty

Typically, we'll want to consider uncertainty when detecting excursions. This could be done with age or paleo ensembles included in the data, but for our synthetic experiment, we'll simulate age and paleo uncertainty. 

```{r}
detectedExcusionEns <- detectExcursion(syntheticExcursion,
                          event.yr = 4200,
                          event.window = 400,
                          ref.window = 400,
                          sig.num = 2,
                          exc.type = "positive",
                          n.ens = 50, #need to test excursions for a number of ensemble member
                          null.hypothesis.n = 50, #how many null hypothesis members to test
                          simulate.time.uncertainty = TRUE,
                          simulate.paleo.uncertainty = TRUE)
```

Now let's look at the results
```{r, warning=FALSE}
print(detectedExcusionEns)
```

and make a plot

```{r, warning=FALSE}
plot(detectedExcusionEns)
```


