---
title: "neotoma"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{neotoma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,message=FALSE,warning=FALSE}
library(actR)
library(neotoma2)
library(lipdR)
library(magrittr)
library(geoChronR)
library(dplyr)
```

# Using actR with Neotoma data
## Get some data from Neotoma

It is now quite simple to get data directly from Neotoma, using the forthcoming `neotoma2` package, either by searching a site using the `get_sites()` function, or directly with `get_downloads()` if you know the dataset number. 


```{r neotoma, message=FALSE,warning=FALSE}
L <- neotoma2::get_sites(sitename = "Bambili 2") %>% 
  get_downloads() %>% 
  neotoma2lipd()
```

You can then pass that "site" object into `neotoma2lipd()`, part of the `lipdR` package, (>= 0.3.0) (for now at least), to create a LiPD object. 


## Create an ensemble age-depth model using geoChronR

If that Neotoma dataset, now a LiPD object, has chronologic data, we can than run it through Bacon or another algorithm in `geoChronR`.


```{r bacon,message=FALSE,warning=FALSE,results='hide',fig.keep='none',fig.keep="all"}
L <- runBacon(L,
              lab.id.var = 'neotomaChronConrolId',
              age.14c.var = NULL,
              age.14c.uncertainty.var = NULL,
              age.var = 'age',
              age.uncertainty.var = 'ageUnc',
              depth.var = 'depth',
              reservoir.age.14c.var = NULL,
              reservoir.age.14c.uncertainty.var = NULL,
              rejected.ages.var = NULL,
              accept.suggestions = TRUE,
              plot.pdf = FALSE)
```

Let's take a look at the age model we created.

```{r plotChron, fig.retina=TRUE}

plotChronEns(L)

```



