---
title: "neotoma"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{neotoma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,message=FALSE,warning=FALSE}
library(actR)
library(neotoma2)
library(lipdR)
library(tidyverse)
library(geoChronR)
```

# Get some data from Neotoma

It is now quite simple to get data directly from Neotoma, using the forthcoming `neotoma2` package, either by searching a site using the `get_sites()` function, or directly with `get_downloads()` if you know the dataset number. 


```{r neotoma, message=FALSE,warning=FALSE}
L <- neotoma2::get_sites(sitename = "Penegor Lake") %>% 
  get_downloads() %>% 
  neotoma2lipd()
```

You can then pass that "site" object into `neotoma2lipd()`, part of the `lipdR` package, (>= 0.3.0) (for now at least), to create a LiPD object. 


If that Neotoma dataset, now a LiPD object, has chronologic data, we can than run it through Bacon or another algorithm in `geoChronR`.


```{r bacon,message=FALSE,warning=FALSE,results='hide',fig.keep='none',fig.keep="all"}
L <- runBacon(L,
              bacon.dir = "~/Dropbox/MacBacon/",
              chron.num = 1,
              meas.table.num = 1,
              lab.id.var = "neotomachronconrolid",
              reservoir.age.14c.var = NULL, 
              reservoir.age.14c.uncertainty.var = NULL, 
              rejected.ages.var = NULL, 
              ask = FALSE,
              plot.pdf = FALSE)
```

Let's take a look at the age model we created.

```{r plotChron, fig.retina=TRUE}

plotChronEns(L)

```



